# Mixed Effects Models

Observations of data aren't always independent, violating an assumption of our GLMs. Never fear, there are statistical models that deal with that; namely, our aforementioned mixed effects models. 

Let's first cite some examples of non-independent observations:
*Time-series data, in which we are resampling the same points multiple times. 
*Blocked data. For example, plants within a plot. We might expect individuals within 1 plot to respond similarly given higher likelihood of relatedness or greater similarity in microsite conditions, relative to plants in other plots.
*Nested data. For example, multiple blood samples taken from an individual. We'd want to indicate that these data are non-independent and likely to be more similar to each other than to samples taken from another individual. 

*Why call these models 'mixed effects'?* 
Because these models contain both fixed and random effects. 
**Fixed effects** = Variables of interest. We are interested in how these factors affect our response variable. Factors and levels of factors have been specifically chosen, and are replicable in future experiments.
**Random effects** = Variables that may explain variation, but that we are not interested in and that we are not interested in replicating in future experiments. For instance, we don't care if we have the same exact block or we sample blood from the same exact person or we collect data in the same year. 

## Examples in R

Let's take a look at running mixed effects models in R. Since agricultural researchers are the royals of blocking, we will take a look at the oat dataset. 

Be sure to load the lmerTest package, since this package will actually generate p-values. Want to read some snarky statistics back and forth?
Check out commentary on developers of the lmer package on why they don't include p-values... sigh...

```{r}
#load libraries and datasets
suppressWarnings(library('faraway'))
suppressWarnings(library('lmerTest'))
suppressWarnings(library('lme4'))
suppressWarnings(library('tidyverse'))
suppressWarnings(library('emmeans'))
suppressWarnings(library('multcomp'))
data('oatvar', package='faraway')
```

First, let's talk about oats. The oats dataset 'Data from an experiment to compare 8 varieties of oats. The growing area was heterogeneous and so was grouped into 5 blocks. Each variety was sown once within each block and the yield in grams per 16ft row was recorded.' - Official description see ?faraway::oats

When we are blocking, I find it really helpful to draw out the statistical design (*Draw out what is happening*). For now, let's just take a look at the data.

```{r}
ggplot(oatvar, aes(y=yield, x=block, color=variety)) + 
    geom_point() +
    geom_line(aes(x=as.integer(block)))
```

Let's just ask how does variety affect yield. The oat-folks have planted the varieties in 5 replicate blocks. This is a smart thing to do, since variation in microsite conditions will cause differences in yield, and they want their yield estimates to reflect this. While microsite effects are interesting, keep in mind, this is not the variable of interest and they would not try to precisely place the same plots. In other words, the blocks are a random variable. We would expect individuals within blocks to respond more similarly to each other than to those plants outside and individuals block, so let's account for that variation! 

The syntax is simple. model <- lmer(responsevariable ~ predictorvariable + (1|block), data=dataset). 

```{r}
oatmodel.1 <- lmer( yield ~ variety + (1|block), data=oatvar)
anova(oatmodel.1)

```
This equation basically says let yield means vary by block. There is a statistically significant effect of variety. So let's do a post-hoc test to see which varieties are different from the other.

```{r}
oatmodel.1 <- lmerTest::lmer( yield ~ variety + (1|block), data=oatvar)
dispairwise1 <- emmeans::emmeans(oatmodel.1, ~variety, type = "response"); dispairwise1
dis_glht <- glht(oatmodel.1, mcp(variety = "Tukey"))

#need to fix the below line
#LetterResults <- multcomp::cld(dis_glht, alpha=0.05, Letters = LETTERS)
```

Add content on nested designs - need to find a decent dataset
```{r}
#why can't i use the oat dataset from MASS?
#model <- lmer(yield ~ variety + Fertilizer + (1|plot/subplot), data=oats)

```

Add content crossed design

Add content random slope and intercept
